---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## safejoin

The package *safejoin* features wrappers around *dplyr*'s functions to join 
safely using various checks

Install package with:

```{r, eval = FALSE}
# install.packages(devtools)
devtools::install_github("moodymudskipper/safejoin")
```

Joining operations often come with tests, one might want to check that:

 1. `by` columns are given explicitly (*dplyr* displays a message if 
   they're not)
 2. Factor columns used for the join have the same levels (*dplyr* displays a
   warning if they don't)
 3. No columns are repeated in both data.frames apart from `by` columns
   (*dplyr* keeps them both and suffixes them silently)
 4. Join columns form a unique key on both or either tables
 5. All rows of both or either tables will be matched
 6. All combinations of values of join columns are present on both or either sides
 
This package provides the possibility for any of these cases to ignore, inform,
warn or abort. 

These checks are handled by a single string parameter, i.e. a sequence of
characters where uppercase letters trigger failures, lower case letters trigger
warnings, and letters prefixed with `~` trigger messages, the codes are as follow:

* `"c"` to check *c*onflicts of *c*olumns
* `"b"` like *"by"* checks if `by` parameter was given explicitly
* `"u"` like *unique* to check that the join colums from an unique key on `x`
* `"v"` to check that the join colums from an unique key on `y`
* `"m"` like *match* to check that all rows of `x` have a _match_
* `"n"` to check that all rows of `y` have a _match_
* `"e"` like *expand* to check that all combinations of joining columns are present in `x`
* `"f"`  to check that all combinations of joining columns are present in `y`
* `"l"`  like *levels* to check that matching is consistent with levels of factor columns

For example, `check = "MN"` will ensure that all rows of both tables are matched.

Additionally when identically named columns are present on both
sides, we can aggregate them into one in flexible ways (including coalesce or
just keeping one of them). This is done through the `conflict` parameter.

The package features functions `safe_left_join`, `safe_right_join`, 
`safe_inner_join`,  `safe_full_join`, `safe_nest_join`,  `safe_semi_join`, 
`safe_anti_join`, and `eat`.

The additional function, `eat`
is designed to be an improved left join in the cases where one is growing a 
data frame. In addition to the features above :

* It uses the `...` argument to select columns from `y` and leverages the select helpers from *dplyr*, allowing also things like renaming on the fly, negative selection, 
  quasi-quotation...
* It can prefix new columns or rename them in a flexible way
* It can summarize `y` on the fly along joining columns for more concise and
readable code

## safe_left_join

*safejoin* offers the same features for all `safe_*_join` functions so we'll
only review `safe_left_join` here, we also limit ourselves to checks of the form
`~*`

We'll use *dplyr*'s data sets `band_members` and `band_instruments` along with
extended versions.

```{r}
library(safejoin)
library(dplyr,quietly = TRUE,warn.conflicts = FALSE)
band_members_extended <- band_members %>%
  mutate(cooks = factor(c("pasta","pizza","spaghetti"),
                        levels = c("pasta","pizza","spaghetti"))) %>%
  add_row(name = "John",band = "The Who", cooks = "pizza")

band_instruments_extended <- band_instruments %>%
  mutate(cooks = factor(c("pizza","pasta","pizza")))

band_members
band_instruments
band_members_extended
band_instruments_extended
```

Not applying any check :

```{r}
safe_left_join(band_members,
               band_instruments,
               check ="")
```

Displaying "Joining, by..." like in default *dplyr* behavior:

```{r}
safe_left_join(band_members,
               band_instruments,
               check ="~b")
```

Check column conflict when joining extended datasets by name:

```{r}
try(safe_left_join(band_members_extended,
                   band_instruments_extended,
                   by = "name",
                   check = "~c"))
```

Check if `x` has unmatched combinations:

```{r}
safe_left_join(band_members_extended,
               band_instruments_extended,
               by = c("name","cooks"),
               check = "~m")
```

Check if `y` has unmatched combinations:

```{r}
safe_left_join(band_members_extended,
               band_instruments_extended,
               by = c("name","cooks"),
               check = "~n")
```

Check if `x` has absent combinations:

```{r}
safe_left_join(band_members_extended,
               band_instruments_extended,
               by=c("name","cooks"),
               check ="~e")
```

Check if `y` has absent combinations:

```{r}
safe_left_join(band_members_extended,
               band_instruments_extended,
               by=c("name","cooks"),
               check ="~f")
```

Check if `x` is unique on joining columns:

```{r}
safe_left_join(band_members_extended,
               band_instruments_extended,
               by=c("name","cooks"),
               check ="~u")
```

Check if `y` is unique on joining columns (it is):

```{r}
safe_left_join(band_members_extended,
               band_instruments_extended,
               by = c("name","cooks"),
               check = "~v")
```

Check if levels are compatible betweeb joining columns:

```{r}
safe_left_join(band_members_extended,
               band_instruments_extended,
               by = c("name","cooks"),
               check = "~l")
```

In case of confict choose either the column from `x` or from `y`:

```{r}
band_members_extended %>% safe_left_join(band_instruments_extended, by = "name", conflict = ~.x)
band_members_extended %>% safe_left_join(band_instruments_extended, by = "name", conflict = ~.y)
```

Our apply any transformation :

```{r}
band_members_extended %>% safe_left_join(band_instruments_extended, by = "name", conflict = coalesce)
band_members_extended %>% safe_left_join(band_instruments_extended, by = "name", conflict = ~coalesce(.y,.x))
band_members_extended %>% safe_left_join(band_instruments_extended, by = "name", conflict = paste)
```

Some common use cases for numerics would be ``confict = `+` ``, `confict = pmin`,
, `confict = pmax`, `confict = ~(.x+.y)/2`.

## eat

All the checks above are still relevant for `eat`, we'll silence them below to
focus on the additional features.

Same as `safe_left_join` :

```{r eat1}
band_members_extended %>% eat(band_instruments_extended)
band_members_extended %>% eat(band_instruments_extended, .by="name", .check = "")
```

The names of `eat`'s parameters start with a dot to minimize the risk of conflict
when naming the arguments fed to the `...`. The `...` are usually used to pass
columns to be eaten, but they are passed to `select` so more features are
available.

Rename eaten columns :

```{r}
band_members_extended %>% eat(band_instruments_extended, .prefix = "NEW", .check = "")
band_members_extended %>% eat(band_instruments_extended, plays, .by="name", .check = "")
band_members_extended %>% eat(band_instruments_extended, -cooks, .by="name", .check = "")
band_members_extended %>% eat(band_instruments_extended, PLAYS = plays, .check = "")
band_members_extended %>% eat(band_instruments_extended, starts_with("p"), .by="name", .check ="")
```

We can check if the dot argument was used by using the character `"d"` in the check string:

```{r}
band_members_extended %>% eat(band_instruments_extended, .check ="~d")
```

In cases of matching to many (i.e. the join columns don't form a unique key for
`y`), we can use the parameter `.agg` to aggregate them on the fly:

```{r}
band_instruments_extended %>% eat(band_members_extended, .check = "")
band_instruments_extended %>% eat(band_members_extended, .agg = ~paste(.,collapse="/"), .check = "")
```

Finally, `conflict = "patch"` is a special value where matches found in `y`
overwrite the values in `x`, and other values are kept. It's different from
`conflict = ~coalesce(.y,.x)` because some values in `x` might be overwritten
by `NA`.

```{r}
band_members_extended %>% eat(band_instruments_extended, .by="name", .conflict = "patch", .check = "")
```
